Hijacking Simulators with Universal Probabilistic Programming

### 2019-08-07 paper/learning/notes.tex

\section{Introduction}
\label{sec:intro}

One challenge in performing inference in such models is the combinatorial increase in \emph{paths}
that can be taken within the given probabilistic model.
This is because each member of the population has 
a bounded, but large number of decisions to make. 
In addition to this, each member of the population can interact
with other members of the population, which presents challenges for current inference methods.
Because in many simulated settings we cannot correctly construct the functional form of the density of the 
desired program. 
This has spawned a large body of research in recent years to understand how to perform 
inference in likelihood free methods~\cite{greenberg2019automatic} \textbf{Add normalising flow citations}
for conditional density estimation. However, whilst many of these methods look to generate a surragote 
for the entire simulator, we propose amortizing only computational expensive sub-components 
of the simulator, which we can do in an automated way with probabilistic programming. 
This enables three things. 
1) It reduces the loss of information, as when simulators are amortized with in current
settings we require, potentially infinite samples, to provide a useful learnt representation of the simulator. 
2) It increases computational efficiency, whilst maintaining expressivity. 
3) It converts expensive rejection sampling loops into to learnt functions that can be 
easily sampled from and used to construct a functional form of the likelihood, that enables the 
target density to be fully constructed, enabling one to leverage a large class of sample effient inference schemes, namely MCMC.

@@

\begin{figure}
  % \begin{minipage}[c]{0.67\textwidth}
  \includegraphics[]{amoritized_rejection.pdf}
  % \end{minipage}\hfill
  % \begin{minipage}[c]{0.3\textwidth}
    \caption{Using PyProb we can locate all regions of inefficient sampling 
    and replace those points in the code with learnt functions that represent
    the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
    by learning surrogate functions for each of the rejection sampling loops.
    In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
    rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
    us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.}
    \label{fig:amortized_sampling}
  % \end{minipage}
\end{figure}

@@

Here the rejection sampling functions take an additional argument $1$, or $0$,
to specify whether or not we want to collect samples to train, $1$, else replace with 
the learnt function, $0$. $P$ represents the program density and $q_{1}$ and $q_{2}$ represents 
the unknown densities that we wish to learn through an amortized approach.
It is important to note that the expected running time of rejection sampler one, $R1$, 
has an expected running time $\mathbb{E}(\mathcal{O}(R1)) = \infty$. 
The simulated densities for both each rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
\begin{figure}
  \centering
  \includegraphics{rs_plots.png}
  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
  sampling block 2, \emph{orange}.}
  \label{fig:rs_simulated}
\end{figure}

@@

@@

\subsection{2nd Baseline}
Instead of using a single Gaussian, use a proposal that is a mixture of Gaussians. 
That is, for $R1$,   $q(z_{1} | z_{2}) = \sum^{N}_{i=1}K_{i}\mathcal{N}(\hat{\mu^{i}_{r1}},  \hat{\sigma^{i}_{r1}}^{2})$
under the constraint that $\sum^{N}_{i=1}K_{i} = 1$. Maybe $N$ can be treated as a hyper-parameter. 
\textbf{TODO: Add math tomorrow}
\subsection{More complicated: Conditional Density estimation}

>>>>

\section{Introduction}
\label{sec:intro}
Simulators arise in a myriad of scientific and industrial settings; epidemiology, physics, engineering, climate modelling and so forth [add citations].
One interesting class of simulators are population-based simulators, using in epidemiology, reinforcement learning
and financial modelling [add citations].
Performing inference in such population-based simulators is challenging, because we have nested-programs, 
which come in the form of rejection sampling blocks. 
In addition to this, each population member can traverse a number of different paths and 
can interact with other members, in certain instances, which can lead to a combinatorial number of paths 
to explore. 
Equally, in many simulated settings we cannot correctly construct the functional form of the density of the desired program,
since we do not have explicit access to the density of the \emph{program}.
Where program density refers to the joint density constructed by the program, which in this case is represented by the simulator.
This inability to evaluate the joint density has spawned research into how to perform 
inference in this likelihood free setting~\cite{greenberg2019automatic} \textbf{Add normalising flow citations}
for conditional density estimation. However, whilst many of these methods look to generate a surragote 
for the entire simulator, we propose amortizing only computational expensive sub-components 
of the simulator, which we can do in an automated way with probabilistic programming. 
This enables three things. 
1) It reduces the loss of information, as when simulators are amortized with in current
settings we require, potentially infinite samples, to provide a useful learnt representation of the simulator. 
2) It increases computational efficiency, whilst maintaining expressivity. 
3) It converts expensive rejection sampling loops into to learnt functions that can be 
easily sampled from and used to construct a functional form of the likelihood, that enables the 
target density to be fully constructed, enabling one to leverage a large class of sample effient inference schemes, namely MCMC.

@@

\begin{figure}
  % \begin{minipage}[c]{0.67\textwidth}
  \includegraphics[width=0.6\textwidth, height=0.6\textheight,keepaspectratio]{amoritized_rejection.pdf}
  % \end{minipage}\hfill
  % \begin{minipage}[c]{0.3\textwidth}
    \caption{Using PyProb we can locate all regions of inefficient sampling 
    and replace those points in the code with learnt functions that represent
    the distributions of the rejection sampling statements. $Z_{i}$ represents the marginal and 
    by learning surrogate functions for each of the rejection sampling loops.
    In doing so we can learn approximate raw sample distributions $f(x_{i} | \eta)$ for each
    rejection sampling loop. This means that we can evaluate the function at each part in the simulator, enabling
    us to construct a target density that can be evaluated when using MCMC inference schemes such as RMH.}
    \label{fig:amortized_sampling}
  % \end{minipage}
\end{figure}

@@

Here the rejection sampling functions take an additional argument $1$, or $0$,
to specify whether or not we want to collect samples to train, $1$, else replace with 
the learnt function, $0$. $P$ represents the program density and $q_{1}$ and $q_{2}$ represents 
the unknown densities that we wish to learn through an amortized approach.
It is important to note that the expected running time of the rejection sampler is expected to geometric mean, we can show this as follows:
\begin{equation}
\text{derivation}
\end{equation}

In addition to this, if we have direct access to a rejection sampling loop then it is indeed 
possible to construct the density, up to a constant of proportionality. 
That is:
\begin{align*}
y \sim q(\theta) \\
do something \\
\alpha = p_{accept}(y,\theta) \\
\mathtt{if} Bern(\alpha) = 1: \\
return y
\end{align*}
thus the density of $y,\theta \propto q(\theta, y) \alpha$.
However, in order to do construct this we need the input $\theta$, $y$, $\alpha$ and the distributions utilised for the rejection sampling loop. 
For any arbitrary simulator, written in a generic code base, this is an incredibly hard 
task to do, because rejection sampling blocks within simulators are not necessarily written in a standard template
\textbf{Add examples in a supplement.  Need to demonstrate to YW that this is challenging to do, even if you know where 
the rejection sampling loops are. Hypothesis, RS loops are not written in a templated form}. With the PPX protocols previously developed in \cite{baydin2018efficient,gram2019hijacking} we can however, 
extract all variables, other than $\alpha$ in an automated way.
However, without being able to easily extract $\alpha$ this means that we are unable to 
correctly build the form of the density and thus we limited in our ability to leverage 
powerful inference methods such as MCMC and Variational inference.
Thus, we propose another approach, automating the amortization of rejection sampling loops. 
To learn functions that we can easily sample from and use to construct the density of the entire
simulator.

The simulated densities for both each rejection sampling block can be seen in Figure~\ref{fig:rs_simulated}.
\begin{figure}
  \centering
  \includegraphics[width=0.2\textwidth, height=0.2\textheight, keepaspectratio]{rs_plots.png}
  \caption{ Simulated densities for rejection sampling block 1, \emph{blue} and rejection
  sampling block 2, \emph{orange}.}
  \label{fig:rs_simulated}
\end{figure}

@@

\subsection{Ground truth}

Because in this simple case we have direct access to $\alpha$, the rejection sampler acceptance probability,
we can calculate the true density up to a constant of proportionality. 
\textbf{I will add this tomorrow}. I had no idea about this, until my meeting with 
YW today.

@@

\subsection{2nd Baseline}
Instead of using a single Gaussian, use a proposal that is a mixture of Gaussians. 
That is, for $R1$,   $q(z_{1} | z_{2}) = \sum^{N}_{i=1}K_{i}\mathcal{N}(\hat{\mu^{i}_{r1}},  \hat{\sigma^{i}_{r1}}^{2})$
under the constraint that $\sum^{N}_{i=1}K_{i} = 1$. Maybe $N$ can be treated as a hyper-parameter. 
Or choose particular values.
\subsection{More complicated: Conditional Density estimation}

\subsection{3rd Baseline}

Use the variational objective constructed above.
